---
title: "Lisa_Wu_HW_2"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(lubridate)
library(arrow)
```

```{r gender}
library(gender)
```

## Load data

Load the following data:
  + applications from `app_data_sample.parquet`
  + edges from `edges_sample.csv`

```{r load-data}
# change to your own path!
data_path <- "/Users/lisawu/Downloads/"
applications <- read_parquet(paste0(data_path,"app_data_sample.parquet"))

applications
```

## Get gender for examiners

We'll get gender based on the first name of the examiner, which is recorded in the field `examiner_name_first`. We'll use library `gender` for that, relying on a modified version of their own [example](https://cran.r-project.org/web/packages/gender/vignettes/predicting-gender.html).

Note that there are over 2 million records in the applications table -- that's because there are many records for each examiner, as many as the number of applications that examiner worked on during this time frame. Our first step therefore is to get all *unique* names in a separate list `examiner_names`. We will then guess gender for each one and will join this table back to the original dataset. So, let's get names without repetition:

```{r gender-1}
library(gender)
#install_genderdata_package() # only run this line the first time you use the package, to get data for it

# get a list of first names without repetitions
examiner_names <- applications %>% 
  distinct(examiner_name_first)

examiner_names
```

Now let's use function `gender()` as shown in the example for the package to attach a gender and probability to each name and put the results into the table `examiner_names_gender`

```{r gender-2}
# get a table of names and gender
examiner_names_gender <- examiner_names %>% 
  do(results = gender(.$examiner_name_first, method = "ssa")) %>% 
  unnest(cols = c(results), keep_empty = TRUE) %>% 
  select(
    examiner_name_first = name,
    gender,
    proportion_female
  )

examiner_names_gender
```

Finally, let's join that table back to our original applications data and discard the temporary tables we have just created to reduce clutter in our environment.

```{r gender-3}
# remove extra colums from the gender table
examiner_names_gender <- examiner_names_gender %>% 
  select(examiner_name_first, gender)

# joining gender back to the dataset
applications <- applications %>% 
  left_join(examiner_names_gender, by = "examiner_name_first")

# cleaning up
rm(examiner_names)
rm(examiner_names_gender)
gc()

```


## Guess the examiner's race

We'll now use package `wru` to estimate likely race of an examiner. Just like with gender, we'll get a list of unique names first, only now we are using surnames.

```{r race-1}
library(wru)

examiner_surnames <- applications %>% 
  select(surname = examiner_name_last) %>% 
  distinct()

examiner_surnames
```
We'll follow the instructions for the package outlined here [https://github.com/kosukeimai/wru](https://github.com/kosukeimai/wru).

```{r race-2}
examiner_race <- predict_race(voter.file = examiner_surnames, surname.only = T) %>% 
  as_tibble()

examiner_race
```

As you can see, we get probabilities across five broad US Census categories: white, black, Hispanic, Asian and other. (Some of you may correctly point out that Hispanic is not a race category in the US Census, but these are the limitations of this package.)

Our final step here is to pick the race category that has the highest probability for each last name and then join the table back to the main applications table. See this example for comparing values across columns: [https://www.tidyverse.org/blog/2020/04/dplyr-1-0-0-rowwise/](https://www.tidyverse.org/blog/2020/04/dplyr-1-0-0-rowwise/). And this one for `case_when()` function: [https://dplyr.tidyverse.org/reference/case_when.html](https://dplyr.tidyverse.org/reference/case_when.html).

```{r race-3}
examiner_race <- examiner_race %>% 
  mutate(max_race_p = pmax(pred.asi, pred.bla, pred.his, pred.oth, pred.whi)) %>% 
  mutate(race = case_when(
    max_race_p == pred.asi ~ "Asian",
    max_race_p == pred.bla ~ "black",
    max_race_p == pred.his ~ "Hispanic",
    max_race_p == pred.oth ~ "other",
    max_race_p == pred.whi ~ "white",
    TRUE ~ NA_character_
  ))

examiner_race
```

Let's join the data back to the applications table.

```{r race-4}
# removing extra columns
examiner_race <- examiner_race %>% 
  select(surname,race)

applications <- applications %>% 
  left_join(examiner_race, by = c("examiner_name_last" = "surname"))

rm(examiner_race)
rm(examiner_surnames)
gc()
```


## Examiner's tenure 

To figure out the timespan for which we observe each examiner in the applications data, let's find the first and the last observed date for each examiner. We'll first get examiner IDs and application dates in a separate table, for ease of manipulation. We'll keep examiner ID (the field `examiner_id`), and earliest and latest dates for each application (`filing_date` and `appl_status_date` respectively). We'll use functions in package `lubridate` to work with date and time values.

```{r tenure-1}
library(lubridate) # to work with dates

examiner_dates <- applications %>% 
  select(examiner_id, filing_date, appl_status_date) 

examiner_dates
```

The dates look inconsistent in terms of formatting. Let's make them consistent. We'll create new variables `start_date` and `end_date`.

```{r tenure-2}
examiner_dates <- examiner_dates %>% 
  mutate(start_date = ymd(filing_date), end_date = as_date(dmy_hms(appl_status_date)))
```

Let's now identify the earliest and the latest date for each examiner and calculate the difference in days, which is their tenure in the organization.

```{r tenure-3}
examiner_dates <- examiner_dates %>% 
  group_by(examiner_id) %>% 
  summarise(
    earliest_date = min(start_date, na.rm = TRUE), 
    latest_date = max(end_date, na.rm = TRUE),
    tenure_days = interval(earliest_date, latest_date) %/% days(1)
    ) %>% 
  filter(year(latest_date)<2018)

examiner_dates
```

Joining back to the applications data.

```{r tenure-4}
applications <- applications %>% 
  left_join(examiner_dates, by = "examiner_id")

rm(examiner_dates)
gc()
```

## Distribution of gender 
```{r gender plotting}
ggplot(data=applications)+
geom_bar(mapping = aes(x = gender), fill = "lightblue")+
  ggtitle("Gender Distribution")+
  xlab("Gender")+
  ylab("Frequency")

```

## Distribution of race
```{r race plotting}
options(scipen=999)
ggplot(data=applications)+ 
geom_bar(mapping = aes(x = race), fill = "pink")+
  ggtitle("Race Distribution")+
  xlab("Race")+
  ylab("Frequency")
```

## Distribution of tenure 
```{r tenure plotting}
ggplot(data=applications)+ 
geom_histogram(mapping = aes(x = tenure_days),width=300, fill = "lightgreen")+
  ggtitle("Tenure Distribution")+
  xlab("Tenure")+
  ylab("Frequency")

```

## Distribution over TCs 
```{r tc plotting}
ttc <- applications$tc
labels <- unique(ttc)
x_labels <- paste0("TC", labels)

ggplot(data=applications) + 
  geom_bar(mapping = aes(x = tc),fill = "orange") +
  scale_x_continuous(breaks = labels, labels = x_labels) +
  ggtitle("Technology Centres Distribution") +
  xlab("Technology centre") +
  ylab("Frequency")

```

## Distribution over WGs 
```{r wg plotting}
ggplot(data=applications)+ 
geom_histogram(mapping = aes(x = examiner_art_unit),binwidth=10, fill = "red")+
  ggtitle("Workgroups Distribution")+
  xlab("Workgroups")+
  ylab("Frequency")
```

## Correlate gender, race with tenure
```{r model 1}
model <- lm(tenure_days ~ 1+ factor(gender) + factor(race), data = applications)
summary(model)
```

## Correlate gender, race, and TC with tenure
```{r model 2}
model_2 <- lm(tenure_days ~ 1+ factor(gender) + factor(race) + factor(tc), data = applications)
summary(model_2)
```

## Findings
This organization has a higher ratio of males than females, and a greater number of people identifying as white than other racial backgrounds. Two models both showed a negative correlation between being a male and tenure days, indicating that increasing one male, the tenure days would decrease. However, coefficient for male changed drastically when including teaching center as a predictor, and the model also shows a better adjusted R-squared. This further confirms that there is a negative correlation between male and tenure days; however, it's not as 131 days but more likely to be a much lower number. The key change when adding TC as a predictor is the correlation between the race white changed from positive to negative. This indicates that being white can reduce tenure days not to increase the days.  